---
title: "Eddie_Cho_BIOMI_609_SPRING_2022_Midterm_Part_2"
output: html_notebook
---
##As Part 2 of the midterm assignment, the Burrow Wheeler Transform (BWT) algorithm & the Suffix Array(SA) algorithm were implemented so that the codes would take a given FASTA input file then produces BWT & SA output files as plain text files

##Please note that the BWT algorithm was loosely based on codes by Geert van Geest (https://rpubs.com/geest007/BWT)

#Step-1: Read a given FASTA file, referencegenome.fasta in this case
```{r}
#1-A: Read lines of a given FASTA file
  file_lines <- readLines("referencegenome.fasta")
#1-B: Remove the first line of the file so that only the strings of A,C,G,T sequences are read
  p_data <-paste(file_lines[2:length(file_lines)], collapse = "")
  print(p_data)
```
#Step-2: Perform BWT & SA then save outputs as .txt
```{r}
#2-A: Create a function called get_output
  get_output <- function(p_data){
    p_data_list <- strsplit(p_data, "")[[1]]  #split strings into list
    p_data_len_list <- length(p_data_list)
    bwm <- matrix(ncol = p_data_len_list, nrow = p_data_len_list)  #empty matrix
    bwm[1,1:p_data_len_list] <- p_data_list  #keep the first row as the same
  
#2-B: Create for loop
  for (i in seq(p_data_len_list-1)){
    p_data_list <- c(p_data_list[p_data_len_list] , p_data_list[1:p_data_len_list-1])
    bwm[i+1, 1:p_data_len_list] <- p_data_list
  }

#2-C: Print BWT Matrix
  print("BWT matrix")
  print(bwm)

#2-D: Sort matrix based on first column along with concatenating the whole rows in the first column
  sorted_index <- order(apply(bwm, 1, function(x) paste(x, collapse = "")))
  print(sorted_index)
  bwm_sorted <- bwm

#2-E: Create for loop 
  for (i in seq(p_data_len_list)){
    row_index <- sorted_index[i]
    bwm_sorted[i, 1:p_data_len_list] <- bwm[row_index, 1:p_data_len_list]
  }

#2-F: print sorted the matrix
  print("Sorted BWT Matrix")
  print(bwm_sorted)

#2-G obtain the last column of sorted matrix
  last_col <- bwm_sorted[1:p_data_len_list,p_data_len_list]
  last_col <- paste(last_col, collapse = "")

#2-H: Obtain the first column for suffix array (lexicographically aka alphabetically sorted)
  first_col <- bwm_sorted[1:p_data_len_list,1]
  first_col <- paste(first_col, collapse = "")

#2-I: Save outputs to text files
  bwt_file<-file("BWT_FASTA_Data.txt")
  writeLines(c(last_col), bwt_file)
  close(bwt_file)
  
  sa_file<-file("Suffix_Array_FASTA_Data.txt")
  writeLines(c(first_col), sa_file)
  close(sa_file)
  
  return(list(last_col, first_col))
}

  output = get_output(p_data)
  bwt = output[[1]]
  sa = output[[2]]

  print("Results of BWT:")
  bwt
  print("Results of SA:")
  sa
```
