---
title: "Eddie_Cho_BIOMI_609_SPRING_2022_Midterm_Part_2"
output: html_notebook
---
##As Part 2 of the midterm assignment, the Burrow Wheeler Transform (BWT) algorithm & the Suffix Array(SA) algorithm were implemented so that the codes would take a given FASTA input file then produces BWT & SA output files as plain text files

##Please note that the BWT algorithm was loosely based on codes by Geert van Geest (https://rpubs.com/geest007/BWT)


## Summary of the Key Sections:
    #1. Reading and Preparing Data:
        #The readLines function reads the FASTA file and extracts only the sequence part (excluding the header line).
    #2. BWT Computation:
        #The code constructs the Burrows-Wheeler Matrix (BWM) by rotating the string in each row and sorting it
         lexicographically to obtain the BWT and Suffix Array.
    #3. Saving and Printing Results:
        #After generating the BWT and Suffix Array results, the code saves them to text files and prints them to the console.

#Step-1: Read a given FASTA file, referencegenome.fasta in this case
```{r}
#1-A: Read lines of a given FASTA file
  file_lines <- readLines("referencegenome.fasta")

#1-B: Remove the first line of the file so that only the strings of A,C,G,T sequences are read
  p_data <-paste(file_lines[2:length(file_lines)], collapse = "") #Combine all sequence lines into one continuous string
  print(p_data)
```

#Step-2: Perform BWT & SA then save outputs as .txt
```{r}
#2-A: Create a function called get_output
  get_output <- function(p_data){
    p_data_list <- strsplit(p_data, "")[[1]]  #split strings into list
    p_data_len_list <- length(p_data_list) #Store the length of the sequence data (number of characters)
    bwm <- matrix(ncol = p_data_len_list, nrow = p_data_len_list)  #empty matrix to store the Burrows-Wheeler matrix
    bwm[1,1:p_data_len_list] <- p_data_list  #keep the first row as the same
  
#2-B: Create for loop
  for (i in seq(p_data_len_list-1)){
    p_data_list <- c(p_data_list[p_data_len_list] , p_data_list[1:p_data_len_list-1]) #Perform the rotation
    bwm[i+1, 1:p_data_len_list] <- p_data_list #Add the rotated sequence as the next row in the BWT matrix
  }

#2-C: Print BWT Matrix
  print("BWT matrix")
  print(bwm) #Output the Burrows-Wheeler matrix

#2-D: Sort matrix based on first column along with concatenating the whole rows in the first column
  sorted_index <- order(apply(bwm, 1, function(x) paste(x, collapse = ""))) #Sort by lexicographical order of rows
  print(sorted_index) #Print the sorted indices to the console
  bwm_sorted <- bwm #Create a copy of the BWT matrix to work with sorted rows

#2-E: Create for loop 
  for (i in seq(p_data_len_list)){
    row_index <- sorted_index[i] #Get the index of the sorted row
    bwm_sorted[i, 1:p_data_len_list] <- bwm[row_index, 1:p_data_len_list] #Reorder rows according to sorted indices
  }

#2-F: print sorted the matrix
  print("Sorted BWT Matrix")
  print(bwm_sorted) #Output the sorted BWT matrix

#2-G obtain the last column of sorted matrix
  last_col <- bwm_sorted[1:p_data_len_list,p_data_len_list] #Extract the last column of the matrix (which is the BWT result)
  last_col <- paste(last_col, collapse = "")  #Collapse the last column into a single string

#2-H: Obtain the first column for suffix array (lexicographically aka alphabetically sorted)
  first_col <- bwm_sorted[1:p_data_len_list,1] #Extract the first column (which is the Suffix Array result)
  first_col <- paste(first_col, collapse = "") #Collapse the first column into a single string

#2-I: Save outputs to text files
  bwt_file<-file("BWT_FASTA_Data.txt") #Open a file for writing the BWT result
  writeLines(c(last_col), bwt_file) #Write the BWT result to the file
  close(bwt_file) #Close the BWT file
  
  sa_file<-file("Suffix_Array_FASTA_Data.txt") #Open a file for writing the Suffix Array result
  writeLines(c(first_col), sa_file) #Write the Suffix Array result to the file
  close(sa_file) #Close the Suffix Array file
  
  return(list(last_col, first_col)) #Return the BWT and Suffix Array results as a list
}

  output = get_output(p_data) #Call the get_output function with the sequence data
  bwt = output[[1]]  #Extract the BWT result from the output list
  sa = output[[2]]  #Extract the Suffix Array result from the output list

  print("Results of BWT:") #Print a label indicating the BWT results
  bwt  #Output the BWT result to the console
  print("Results of SA:") #Print a label indicating the Suffix Array results
  sa #Output the Suffix Array result to the console
```
